---
title: "Data Analysis of Storms Across the USA"
author: "Christiano Weirich"
date: '2022-09-09'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis
***
The [National Oceanic and Atmospheric Administration](https://www.noaa.gov/) is
an U.S. government agency that works with data related to various weather 
and other natural phenomena with the objective of making forecasts and other
analysis. The purpose of this report is to use their [Storm Database](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)
to answer two questions:

1. Which types of events are most harmful with respect to population health?

2. Which types of events have the greatest economic consequences?

## Data Processing
***
The *dplyr* library was used to process the data
```{r message=F}
library(dplyr)
```

And the environment in the moment of the analysis was
```{r}
sessionInfo()
```

The data was downloaded as of this document creation date from the URL mentioned 
in the Synopsis above and was read into memory
```{r cache=T}
rawData <- read.csv("repdata_data_StormData.csv.bz2")
```

## Results
***
### Which types of events are most harmful with respect to population health?
***
To address the first question, the raw data was grouped by **EVENT TYPE** and 
summarized to show the sum of **FATALITIES** and **INJURIES** ranked in 
descendant order
```{r}
healthData <- rawData %>% 
                group_by(EVTYPE) %>% 
                summarise(
                    FATALITIES = sum(FATALITIES),
                    INJURIES = sum(INJURIES)) %>% 
                mutate(TOTAL = FATALITIES + INJURIES) %>%
                arrange(desc(TOTAL))
healthData
```
So these are the most harmful events regarding population health.

```{r fig.width=10}
top5MostHarmful <- healthData[1:5,]
barplot(top5MostHarmful$TOTAL, 
        names.arg = top5MostHarmful$EVTYPE, 
        main="Top 5 Most Harmful Events", 
        ylab="Number of Casualties", 
        col="red")
```
As the plot illustrates, Tornadoes are the most harmful events to public health
by a great margin.

***
### Which types of events have the greatest economic consequences?
***
To subset the data for the second answer, we will again group the data by **EVENT TYPE**
but now use the **PROPERTY DAMAGE** and **CROP DAMAGE** values to be able to
rank the biggest losses in economic terms.
As property and crop values are divided in two columns each, one with the amount
of money and the other with the exponent of magnitude of the amount, a new
column is created with the amount in absolute values.

```{r}
rawData$PROPDMGEXP<-recode(rawData$PROPDMGEXP,
                           'K'=1000,
                           'M'=1000000,
                           'B'=1000000000,
                           .default=1)
rawData$CROPDMGEXP<-recode(rawData$CROPDMGEXP,
                           'K'=1000,
                           'M'=1000000,
                           'B'=1000000000,
                           .default=1)
rawData$ABSPROPDMG <- rawData$PROPDMG*rawData$PROPDMGEXP
rawData$ABSCROPDMG <- rawData$CROPDMG*rawData$CROPDMGEXP
economicData <- rawData %>% 
                group_by(EVTYPE) %>% 
                summarise(
                    ABSPROPDMG = sum(ABSPROPDMG),
                    ABSCROPDMG = sum(ABSCROPDMG)) %>% 
                mutate(TOTAL = ABSPROPDMG + ABSCROPDMG) %>%
                arrange(desc(TOTAL))
economicData
```
Giving us the events causing the most economic losses.

```{r fig.width=10}
top5BiggestLosses <- economicData[1:5,]
barplot(top5BiggestLosses$TOTAL, 
        names.arg = top5BiggestLosses$EVTYPE, 
        main="Top 5 Events Causing Economic Losses", 
        ylab="Losses (in USD)", 
        col="purple")
```
The plot shows that Floods are the most disastrous events in regard of the 
economy.  

## Conclusion
***
As demonstrated by this analysis, **Tornadoes** are the events that cause more
casualties and raises more concerns in the public health area and **Floods** are
the events that cause the biggest economic losses.